Chaotic mixing of tidal debris

### 2015-06-25 ms.tex
redo disc and conc w kvj and sp comments

\section{Discussion}\label{sec:discussion}

\subsection{Existence of streams}

We have considered here a simple triaxial gravitational potential chosen to mimic the median properties of dark matter halos formed in dark-matter-only simulations (or the large-scale properties of halos formed in simulations with baryonic effects). We have found that `mean' chaos indicators such as the Lyapunov exponent or frequency diffusion rate predict that chaos should be dynamically unimportant in the halos of galaxies for the majority of orbits (e.g., the predicted timescales are 100-1000s of times longer than the typical orbital times, at least in this simple potential). However, while the mean chaotic evolution of, for example, the fundamental frequencies of an orbit may be prohibitively long, the frequencies of even mildly chaotic orbits do evolve on a smaller scale over much shorter times. The magnitude of these variations depends on the resonant structure local to the orbit (which is highly potential-dependent), but this result is generic to any potential with resonances and regions of stochasticity. When this small-scale but short-time frequency evolution has an amplitude comparable to or larger than the intrinsic spread in orbital frequencies for tidal debris, the debris will appear dynamically hotter and become diffuse faster than along a regular orbit. We have found that even in the simple potential used in this work (Section~\ref{sec:potential}), a large fraction of orbits would naively be considered weakly chaotic, but tidal streams formed on these orbits would appear morphologically different (e.g., more diffuse) than those formed on regular orbits.

@@

In this work, we have only considered low-mass progenitor systems such as globular clusters because the intrinsic spreads in fundamental frequencies are small (0.1--0.5\%). Small changes to the frequencies of the orbits of tidal debris disrupted from these progenitors due to chaotic processes will therefore cause observable changes to the real-space morphology of the debris. For more massive progenitor systems, the typical size and velocity dispersion of the debris will be larger and thus the debris morphology will be less sensitive to small changes in orbital properties. The mass-scale of the debris that will display enhanced density evolution depends on the magnitude of weak chaos, which depends on the orbit structure of a given potential. In potentials with more significant chaos, debris disrupted from more massive progenitor systems may also display stream-fanning.

The mass of the progenitor sets the physical scale of the debris as it is disrupted. For more massive progenitors and therefore larger spreads in debris, different orbits may have different chaotic diffusion timescales that depend on the details of the disruption (e.g., the orbital phase of the pericenter, or the energy offset of the debris). This could potentially lead to interesting density variations along the tidal debris.

The Ophiuchus stream \citep{bernard14, sesar15} appears to be a thin, short tidal stream (deprojected length $\approx$1.5--2 kpc) near the Galactic bulge with no apparent progenitor. At Galactocentric $R \approx 1$ kpc, $z \approx 5$ kpc, the orbits of the stream stars likely pass through the MW disk, feel the triaxiality of the Galactic bar \citep[e.g.,][]{wegg13, wegg15}, and have short orbital periods (relative to streams in the halo). It is possible that the observed debris is the last remnants of the recently disrupted progenitor system \citep{sesar15}, however if a significant number of stars were disrupted on previous pericentric passages, this older debris may be `fanned' and still near the observed portion of the stream. The fanned debris would be much lower surface brightness and thus much more difficult to detect. However, the detection of such stars would confirm that enhanced density evolution is a dynamically relevant process for thin streams in real galaxies.

\subsection{The dangers of stream modeling}

Tidal stream modeling is one of the most promising ways to constrain the 3D mass distribution around the Milky Way at distances of 10s to 100s of kpc. Methods that use tidal streams to infer properties of the Galactic potential typically operate by constructing models of the debris distribution using either the present-day phase-space density \citep[e.g.,][]{kuepper12, kuepper15, amorisco and others}, time-of-disruption phase-space density \citep{apw13, apw14}, or the density in angle-action coordinates \citep{sanders14, bovy14}. All of these methods may fail or produce uninterpretable results if modeling globular-cluster streams on mildly chaotic orbits.

Consider a globular cluster system that disrupts along a mildly chaotic orbit so that a significant portion of the resulting stream is more diffuse (lower surface brightness) and therefore unobservable due to observational biases. The high-density part of the stream may still appear thin (e.g., the recently disrupted debris in the middle panels of Figure~\ref{fig:nbodysims}). If the stream modeling focuses on reproducing the present-day density of the stream and the potential is restricted to an integrable, analytic form, the phase-space distribution of the debris may be well-reproduced by a progenitor on a particular orbit in such a potential. However, the inferences about the potential or mass distribution may be biased significantly and the inferences about the evolution time and internal properties of the system will certainly be incorrect --- for example, the evolution time may be biased to shorter times to reproduce the apparent shortness of the stream. With realistic observational uncertainties, it is likely the case that for any stream formed from a progenitor on a weakly chaotic orbit, there is a regular progenitor orbit in some other potential that produces a model stream equally consistent with the data.

Due to the often unknown or poorly understood selection biases involved in kinematic observations of stream stars (due to, e.g., survey limits, dust extinction, and stellar population variations), modeling the final phase-space densities of tidal streams will lead to uninterpretable inferences about the potential of the host galaxy. We argue that by instead performing the density modeling at the time of disruption \citep[e.g.,][]{apw14}, uncertainties in, for example, the length or width of the stream will not bias potential parameter inferences as significantly (see Price-Whelan et al. 2015, in prep.). In either case, it is critical that potential modeling efforts with tidal streams allow enough freedom in the potential to explore all possible configurations, though this quickly becomes a high-dimensional, multi-modal, and therefore computationally prohibitive inference problem.

\todo{apw}{Need more coherent words about what happens when trying to use angle-action coordinates...}.

The potential considered in this work is `unrealistic' for the total potential of a Milky-Way-like galaxy in that it is static, smooth, and does not contain baryonic components. Hydrodynamic simulations of forming galactic disks in cosmological dark matter halos have shown that baryonic feedback and relaxation can significantly change the inner distribution of dark matter and either make the potential more spherical or oblate \citep{many}. However, the significance of baryonic relaxation or of time-dependence, triaxiality, and substructure on shaping the matter distribution within the Milky Way is largely unknown. We now consider each of these effects independently and comment on the expected consequences of these perturbations.

\subsubsection{Baryonic components}

\cite[][D08]{debattista08} and \cite[][V10]{valluri10} studied the orbit evolution induced by growing a baryonic disk in dark matter halos with various shapes and orientations (e.g., prolate, triaxial). In the simulations of D08, the disk component is artificially fixed along one of the principal axes of the halo potential and the mass of the disk is grown adiabatically. D08 found that (in general) the growth of a disk slowly deforms the orbits of mass tracers (e.g., dark matter particles) and preferentially populates tube orbits but does not result in a large degree of chaotic scattering. Consequently, the inner shape of the potential becomes more oblate or spherical due to the prevalence of tube orbits.

If the inner halos of galaxies are indeed close to spherical or oblate, the majority of orbits will be regular and chaos will be less important, however this is far from conclusive. First, there are observational hints that the intrinsic alignment of galactic disks may be close to random \citep{??} (rather than fixed to symmetry planes of the host halo potential). Simulations since the work of D08 have shown that by `freeing' the disk component, the angular momentum vector of the disk can precess and wander erratically away or around the principal axes of the host halo \citep{debattista15}. It is unknown what effect this would have on the orbit structure of nearby halo orbits, but we naively expect that this would destabilize a significant number of orbits. Second, if the outer potential remains triaxial, the transition region between, e.g., oblate and triaxial may be populated by chaotic orbits. We have found from simple experiments in superposing potential components that transition regions between potential shapes induce significant amounts of chaos. Though superposing potentials is not realistic, it is at least suggestive that such configurations might be unstable. A more careful exploration of this phenomenon will be presented in future work. Lastly, the existence of a bulge or thin disk component may enhance chaos through a process similar to periodic forcing of the orbit. Passage through a thin disk results in a nearly instantaneous, impulsive forcing of the orbit at a frequency set by the circulation frequency of the orbit. This may act to increase the significance of chaos along disk-crossing orbits \citep[e.g.,][]{kandrup00, hunter05}. This effect is likely unresolved in many simulations where the force and spatial resolution is insufficient to capture this effect.

\subsubsection{Subhalos}

Cosmological simulations predict that dark matter halos are filled with substructure in the form of dark matter subhalos. If they exist, these subhalos may account for up to $\approx$1-10\% of the mass of the dark matter \citep{??} and therefore may contribute significantly to and orbit in the large-scale potential of any galaxy. Gravitational scattering due to subhalo interactions has been studied, however in the halos of galaxies where dynamical times are long, the scattering cross-section for \emph{strong} encounters is very small \citep{??}. Instead, the collective effect of the subhalos may instead act as a noise term in the Hamiltonian of any halo orbit. This subhalo-induced colored noise --- which depends on the mass spectrum and distribution of subhalos --- may also act to simply increase the magnitude of chaos along orbits, and destabilize sufficiently non-resonant orbits \citep[see, e.g.,][]{kandrup00}.

\subsubsection{Time-dependence}

Galaxies are certainly not static systems. To first order, galaxies grow in mass --- for example, the spherically averaged mass profile of dark matter halos evolves fairly predictably in cosmological simulations \citep{wechsler02, buist14} after some initial period of stochastic mass growth. The Milky Way has probably had a fairly calm accretion history over the last 6 Gyr and therefore the mass growth may be similar to that seen in simulations. This steady growth most likely does not alter the global structure or shape of the potential. However, we also know from simulations that figure rotation, baryonic feedback, and the accretion and phase-mixing of subhalos do perturb the global state of simulated halos. \cite{deibel11} showed that by adopting pattern speeds comparable to those found in cosmological simulations, figure rotation generally acts to destabilize orbits (rather than stabilize chaotic orbits in the equivalent static potential) and the resulting orbit structure is that most regular orbits are associated with resonances. In future work we will explore the effect of these time-dependent processes on the chaotic dispersal of tidal streams using live potentials from cosmological N-body simulations.

The large-scale shapes of galactic gravitational potentials are thought to be triaxial and therefore contain an appreciable number of chaotic orbits. Chaos is not thought to be dynamically important for shaping the global structure of galaxies because the predicted chaotic timescales such as the Lyapunov or frequency diffusion times for orbits in galactic-like potentials are very long (many 1000s of orbital periods). However, these timescales measure the mean evolution of orbital properties such as the fundamental frequencies and do not capture the full nature of chaotic orbit diffusion.

We have numerically computed the magnitude of chaos for a large grid of iso-energy orbits in a triaxial gravitational potential using two independent methods that have been used extensively to classify and characterize chaotic orbits: the Lyapunov exponent and frequency diffusion rate. From each of these indicators, we compute a timescale over which chaos is likely to be important and find that the majority of orbits have chaotic timescales greater than 100s of orbital periods, however with the frequency diffusion rate we are still able to resolve weak chaos. We then study the density evolution of small ensembles of orbits generated around each orbit in the grid used for the previous experiment and find that along some orbits classified as weakly chaotic --- with chaotic timescales of 100s of orbital periods --- the orbit ensembles display enhanced density evolution and reach a lower overall density faster than orbit ensembles around nearby regular orbits. We propose an explanation for this apparent discrepancy by considering the nature of chaotic diffusion: the classical chaos indicators are most sensitive to the  slow, Arnold diffusion process that can cause large changes to orbital properties, but small-scale frequency evolution occurs over much shorter times as chaotic orbits stochastically diffuse across the stochastic layers that surround many resonances. We propose a new measure of chaos with tidal stream morphological evolution in mind: the amplitude of small-scale, fractional changes in frequencies over a given number of orbital periods along a stream progenitor orbit. When this amplitude is comparable to or larger than the typical spread in orbital properties of tidal debris from the progenitor, the phase-space density of the debris will evolve faster to a state of lower density relative to nearby regular orbits.

\todo{apw}{Concluding paragraph that drives home the importance of this work...}
The timescales in this paper are not meant to be interpreted physically, as the potential is a very crude approximation to the total potential of a Milky-Way-like galaxy. The key point is that even when the mean chaos indicators predict a long chaotic timescale, streams may display enhanced density evolution over much shorter times. Tidal debris is sensitive to the small-scale, short-time evolution of orbital properties (e.g., the fundamental frequencies) induced by weak chaos.

>>>>

\section{Discussion: limitations and future work}\label{sec:discussion}

% Something about this: The timescales in this paper are not meant to be interpreted physically, as the potential is a very crude approximation to the total potential of a Milky-Way-like galaxy. The key point is that even when the mean chaos indicators predict a long chaotic timescale, streams may display enhanced density evolution over much shorter times. Tidal debris is sensitive to the small-scale, short-time evolution of orbital properties (e.g., the fundamental frequencies) induced by weak chaos.

We have shown that the Lyapunov and frequency diffusion times are indicators of chaos and that the frequency diffusion time resolves the detailed resonant structure of gravitational potentials, but the timescales predicted do not capture the importance of chaos for the density evolution of ensembles of orbits meant to mimic tidal debris. We have shown that small-scale but fast chaotic diffusion of orbits can explain this enhanced mixing, and have proposed a new indicator to characterize the importance of this effect over given evolution times. In the sub-sections below, we discuss a few important limitations that remain for exploration in future work.

We have only considered low-mass progenitor systems such as globular clusters because the intrinsic spreads in fundamental frequencies are small (0.1--0.5\%). Small changes to the frequencies of the orbits of tidal debris disrupted from these progenitors due to chaotic processes will therefore cause observable changes to the real-space morphology of the debris. For more massive progenitor systems, the typical size and velocity dispersion of the debris will be larger and thus the debris morphology will be less sensitive to small changes in orbital properties. The mass-scale of the debris that will display enhanced density evolution depends on the magnitude of weak chaos, which depends on the orbit structure of a given potential. In potentials with more significant chaos, debris disrupted from more massive progenitor systems may also display `stream-fanning.'

The potential considered in this work is `unrealistic' for the total potential of a Milky-Way-like galaxy in that it is static, smooth, and does not contain baryonic components. Simulations of forming galactic disks in cosmological dark matter halos have shown that baryonic feedback and relaxation can significantly change the inner distribution of dark matter and either make the potential more spherical or oblate \citep{many}. However, the significance of baryonic relaxation or of time-dependence, triaxiality, and substructure on shaping the matter distribution within the Milky Way is largely unknown. Here we briefly summarize future directions for potential models:
\begin{itemize}
	\item \emph{Baryonic components}: \cite[][D08]{debattista08} and \cite[][V10]{valluri10} studied the orbit evolution induced by growing a baryonic disk in dark matter halos with various shapes and orientations (e.g., prolate, triaxial). In general, the authors find that the growth of a disk slowly deforms the orbits of mass tracers (e.g., dark matter particles) and preferentially populates tube orbits. Consequently, the inner shape of the potential becomes more oblate or spherical due to the prevalence of tube orbits. If the inner halos of galaxies are indeed close to spherical or oblate, the majority of orbits will be regular and chaos will be less important, however this is far from conclusive. We have found from simple experiments in superposing potential components that transition regions between potential shapes can actually enhance significant amounts of chaos. Though superposing potentials is not realistic, it is at least suggests that a more careful exploration of potential configurations is required to understand how complex, radius-dependent potential forms affect the amount and significance of chaos in galactic halos.

\item \emph{Time dependence}: Galaxies are certainly not static systems. To first order, galaxies grow in mass --- for example, the spherically averaged mass profile of dark matter halos evolves fairly predictably in cosmological simulations \citep{wechsler02, buist14} after some initial period of stochastic mass growth. The Milky Way has probably had a fairly calm accretion history over the last 6 Gyr and therefore the mass growth may be similar to that seen in simulations. This steady growth most likely does not alter the global structure or shape of the potential. However, we also know from simulations that figure rotation, baryonic feedback, and the accretion and phase-mixing of subhalos do perturb the global state of simulated halos. \cite{deibel11} showed that by adopting pattern speeds comparable to those found in cosmological simulations, figure rotation generally acts to destabilize orbits (rather than stabilize chaotic orbits in the equivalent static potential) and the resulting orbit structure is that most regular orbits are associated with resonances. In future work we will explore the effect of these time-dependent processes on the chaotic dispersal of tidal streams using live potentials from cosmological N-body simulations.

\item \emph{Substructure}: Cosmological simulations predict that dark matter halos are filled with substructure in the form of dark matter subhalos. If they exist, these subhalos may account for up to $\approx$1-10\% of the mass of the dark matter \citep{??} and therefore may contribute significantly to and orbit in the large-scale potential of any galaxy. Gravitational scattering due to subhalo interactions has been studied, however in the halos of galaxies where dynamical times are long, the scattering cross-section for \emph{strong} encounters is very small \citep{??}. Instead, the collective effect of the subhalos may instead act as a noise term in the Hamiltonian of any halo orbit. This subhalo-induced colored noise --- which depends on the mass spectrum and distribution of subhalos --- may also act to simply increase the magnitude of chaos along orbits, and destabilize sufficiently non-resonant orbits \citep[see, e.g.,][]{kandrup00}.

\end{itemize}

\subsection{Stream modeling}

Tidal stream modeling is one of the most promising ways to constrain the 3D mass distribution around the Milky Way at distances of 10s to 100s of kpc. Methods that use tidal streams to infer properties of the Galactic potential typically operate by constructing models of the debris distribution using either the present-day phase-space density \citep[e.g.,][]{kuepper12, kuepper15, amorisco and others}, time-of-disruption phase-space density \citep{apw13, apw14}, or the density in angle-action coordinates \citep{sanders14, bovy14}. All of these methods may fail or produce uninterpretable results if modeling globular-cluster streams on mildly chaotic orbits. For each of these methods, it is important to understand the failures and biases introduced by ignoring chaotic orbit evolution.

The Ophiuchus stream \citep{bernard14, sesar15} appears to be a thin, short tidal stream (deprojected length $\approx$1.5--2 kpc) near the Galactic bulge with no apparent progenitor. At Galactocentric $R \approx 1$ kpc, $z \approx 5$ kpc, the orbits of the stream stars likely pass through the MW disk, feel the triaxiality of the Galactic bar \citep[e.g.,][]{wegg13, wegg15}, and have short orbital periods (relative to streams in the halo). It is possible that the observed debris is the last remnants of the recently disrupted progenitor system \citep{sesar15}, however if a significant number of stars were disrupted on previous pericentric passages, this older debris may be `fanned' and still near the observed portion of the stream. The fanned debris would be much lower surface brightness and thus much more difficult to detect. The detection of this low-surface-brightness component would open up the possibility that enhanced density evolution due to chaos is a dynamically relevant process for thin streams in real galaxies, though \cite{carlberg15} has shown that it is also possible also to get short, high density segments of streams from streams formed on eccentric orbits.

%The large-scale shapes of galactic gravitational potentials are thought to be triaxial and therefore contain an appreciable number of chaotic orbits. Chaos is not thought to be dynamically important for shaping the global structure of galaxies because the predicted chaotic timescales such as the Lyapunov or frequency diffusion times for orbits in galactic-like potentials are very long (many 1000s of orbital periods). However, these timescales measure the mean evolution of orbital properties such as the fundamental frequencies and do not capture the full nature of chaotic orbit diffusion.

We have considered here a simple triaxial gravitational potential chosen to mimic the median properties of dark matter halos formed in dark-matter-only simulations (or the large-scale properties of halos formed in simulations with baryonic effects). We have numerically computed the magnitude of chaos for a large grid of iso-energy orbits in this potential using two independent methods that have been used extensively to classify and characterize chaotic orbits: the Lyapunov exponent and frequency diffusion rate. From each of these indicators, we compute a timescale over which chaos is likely to be important and find that the majority of orbits have chaotic timescales greater than 100s of orbital periods, however with the frequency diffusion rate we are still able to resolve weak chaos. We then study the density evolution of small ensembles of orbits generated around each orbit in the grid used for the previous experiment and find that along some orbits classified as weakly chaotic --- with chaotic timescales of 100s of orbital periods --- the orbit ensembles display enhanced density evolution and reach a lower overall density faster than orbit ensembles around nearby regular orbits (which mix due to phase-mixing alone). We propose an explanation for this apparent discrepancy by considering the nature of chaotic diffusion: the classical chaos indicators are most sensitive to the slow, Arnold diffusion process that can cause large changes to orbital properties, but small-scale frequency evolution occurs over much shorter times as chaotic orbits stochastically diffuse across the stochastic layers that surround many resonances. We propose a new measure of chaos with tidal stream morphological evolution in mind: the amplitude of small-scale, fractional changes in frequencies over a given number of orbital periods along a stream progenitor orbit. When this amplitude is comparable to or larger than the typical spread in orbital properties of tidal debris from the progenitor, the phase-space density of the debris will evolve faster to a state of lower density relative to nearby regular orbits.

@@

\todo{apw}{Last paragraph that drives home the importance of this work...too tired to finish this tonight...}
